The Strucplot Framework---Visualizing Multi-way Contingency Tables

### 2005-09-18 trunk/vcd/inst/doc/strucplot.Rnw
- vignettes revised - bugs in structable indexing fixed - mosaic, sieve, and assoc made structable aware - panel_generator class -> generating_function - offset_labels and offset_varnames added to labeling_border - spacing_highlighting replaces spacing_doubledecker
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}

@@

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
@

\title{strucplots---Visualizing higher-dimensional Contingency Tables in \pkg{vcd}}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\url{David.Meyer@R-project.org}
}
\maketitle
\sloppy

The \code{strucplot} framework in \pkg{vcd} can be used to visualize
higher-dimensional contingency tables and integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. To allow for more than
two dimensions, the plot area is constructed using recursive
conditional splits, given the tiles representing the levels of the
previous dimension.
This principle defines a graph `family' with still many parameters
left such as:

\begin{itemize}
\item the split direction for each dimension
\item the content of the tiles
\item the graphical parameters of the tiles' content
\item the spacing between the tiles
\item the labeling of the tiles
\end{itemize}

@@

The strucplot framework is highly modularized; Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
The core function is \code{strucplot()} which processes the parameters,
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates panel, labeling, shading, and spacing functions to produce 
the plot. The actual workhorse is the chosen panel function: currently, 
low-level panel functions for association plots (\code{struc\_assoc}), 
mosaic plots (\code{struc\_mosaic}), and sieve plots (\code{struc\_sieve})
are available. In addition, there are
high-level wrapper functions such as \code{mosaic()}, \code{assoc()},
and \code{sieve()} to \code{strucplot()} that conveniently set the 
parameters---it is these high-level functions that the user normally would like to use, 
although \code{strucplot()} is fully functional on its own. \code{mosaic()}, 
\code{assoc}, and \code{sieve()} are generic functions: 
for both, there exist a formula method and a default 
method. Other functions, such as \code{doubledecker()}, \code{pairs()}, and 
\code{cotabplot}, are on even higher level and make use of 
code{mosaic()}, \code{assoc()}, and \code{sieve()} to produce plots.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{struc}
  \caption{Strucplot components.}
  \label{fig:struc}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics{vcdViewport}
  \caption{Viewport layout for strucplots with their names. [A] =
    ``cornerTL'', [B] = ``cornerTR'', [C] = ``cornerBL'', [D] = ``cornerBR''.}
  \label{fig:layout}
\end{figure}

\section{Mosaic and Association Plots}

As an example, consider the famous `Survival on the Titanic' 
data containing two binary variables (survival to the disaster and
gender), as well as two categorical variables (class and age). The
`flattened' contincency table can be obtained using the
\code{structable()} function (quite similar to \code{ftable()}, but
allowing the specification of split directions as in mosaic():

<<structable,results=verbatim>>=
structable(Titanic)
@

Let us first visualize the contingency table by the means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits (see Figure \ref{fig:observed}):

<<Observed>>=
mosaic(Titanic, main = TRUE)
@

\noindent The small bullets indicate zero entries in the corresponding cell.

\begin{figure}[h]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent  It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

<<Expected>>=
mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent In order to directly compare observed and expected values,
we can use a sieve plot (see Figure \ref{fig:sieve}):

<<sieve>>=
sieve(Titanic)
@

\begin{figure}[h]
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE>>=
<<sieve>>
@
\caption{Sieve plot for the Titanic data visualizing simultaneously
  observed and expected values.}
\label{fig:sieve}
\end{center}
\end{figure}

\noindent The next step is to directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
function scales to more than two variables (see Figure \ref{fig:residuals}):

<<Residuals>>=
assoc(Titanic, compress = FALSE)
@

\begin{figure}[h]
\begin{center}
<<Residualsfig,fig=TRUE,echo=FALSE>>=
<<Residuals>>
@
\caption{Association plot for the Titanic data.}
\label{fig:residuals}
\end{center}
\end{figure}

@@

So far, we have visualized the full table. Parts of the data can be
selected either using \code{margin.table()} on the input data, or the
formula interfaces. The following example compares the age
distribution among classes, given gender, and also shows how to 
use grid's viewport framework to put two plots besides 
each other:

pushViewport(viewport(layout.pos.col = 1))
mosaic(~ Class + Age, data = Titanic[,"Male",,], 
       margin = c(bottom = 0, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(~ Class + Age, data = Titanic[,"Female",,], 
       margin = c(bottom = 0, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
<<Variables>>
@
\caption{Age distribution among classes, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent note the use of the \code{margin} argument: it takes a
vector with up to four values whose unnamed components are recycled,
but ``overruled'' by the named arguments. Thus, in the example, only
the bottom margin is set to 0, leaving the default size of 2 lines
unchanged for the other margins. This idea applies to almost all
vectorized arguments in the strucplot framework (with
\code{split\_vertical} as a prominent exception).

Since mosaic displays are ``conditional plots'' by definition, we can
also use one single mosaic for stratified plots.
The formula interface of \code{mosaic()} allows the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

<<Conditioning>>=
mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
@

\begin{figure}[h]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}

\noindent The effect of using this kind of formula is that
conditioning variables are permuted before the conditioned variables
in the table, and that \code{spacing\_conditional()} is used to better
distinguish conditioning from conditioned dimensions.
Note, however, that the plots in the ``pseudo-strata'' are
distorted since they are not corrected for the marginal
distribution(s) of the conditioning variables. Indeed,
the \code{cotabplot()} function would do a much better job here (see
next section).

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one). Sensible recycling rules apply. 
For compatibility with \code{mosaicplot()}, the \code{mosaic()} function
also allows the use of a \code{"direction"} argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figures \ref{fig:split1} and \ref{fig:split2}):

<<split1>>=
mosaic(Titanic, direction = "h")
@

<<split2>>=
mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
@

\begin{figure}[h]
\begin{center}
<<split1fig,fig=TRUE,echo=FALSE>>=
<<split1>>
@
\caption{Mosaic plot for the Titanic data---alternative splitting (1).}
\label{fig:split1}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
<<split2fig,fig=TRUE,echo=FALSE>>=
<<split2>>
@
\caption{Mosaic plot for the Titanic data---alternative splitting (2).}
\label{fig:split2}
\end{center}
\end{figure}

\section{Higher-level functions}

The last plot pretty much resembles a doubledecker plot, except for
the spacing. Since spacing is also modular in the strucplot framework,
the \code{doubledecker()} function was straightforward to implement:
it really is just a wrapper for \code{mosaic()}, setting the right
defaults. Figure \ref{fig:doubledecker} shows a doubledecker plot of
the Titanic data, explaining the probability of Survival by Age, given
Sex, given Class:

@@

\begin{figure}[h]
\begin{center}
<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
<<doubledecker1>>
@
\caption{Doubledecker plot for the Titanic data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

Another high-level function is \code{pairs.table()}, that is, the S3-method for objects
of class "table" of the \code{pairs()} generic. This function produces a matrix 
of strucplots (currently, association or mosaic plots) in the off-diagonal cells, and
the variable names (with, optionally, univariate statistics) in the diagonal cells. Figure
\ref{fig:pairs} shows mosaic plots visualizing mutual independence in the lower triangle,
association plots for the same in the upper triangle, and bar charts in the diagonal.

<<pairs>>=
pairs(Titanic, upper_panel = pairs_assoc)
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairsfig,fig=TRUE,echo=FALSE>>=
<<pairs>>
@
\caption{Pairs plot for the Titanic data.}
\label{fig:pairs}
\end{center}
\end{figure}

\noindent Each cell's row and column define two variables $X$ and $Y$ used for the
specification of four different types of independence: `\code{pairwise}',
     `\code{total}', `\code{conditional}' and `\code{joint}'. The pairwise mosaic
     matrix shows bivariate marginal relations between $X$ and $Y$, collapsed over all
     other variables. The total independence mosaic matrix shows mosaic
     plots for mutual independence, i.e., for marginal and conditional
     independence among all pairs of variables. The conditional
     independence mosaic matrix shows mosaic plots for marginal
     independence of $X$ and $Y$ given all other variables. The joint independence
     mosaic matrix shows mosaic plots for joint independence of all
     pairs ($X$, $Y$) of variables from the others.

Since the matrix is symmetric, the upper and lower part can be used to display different 
types of independence models, or different strucplots (currently, only association and 
mosaic plots are supported). The available panel functions (\code{pairs\_assoc()} and 
\code{pairs\_mosaic()}) are just simple wrappers to \code{assoc()} and 
\code{mosaic()}. Obviously, seeing patterns in strucplot matrices 
becomes increasingly difficult with
higher dimensionality. Typically, therefore, this plot is used with a suitable 
residual-based shading (described in a separate vignette).

Another useful high-level plot is produced by \code{cotabplot()}: this function can be
used for plotting stratified strucplots in a lattice-like display. Whereas \code{pairs()} 
is conditioning on variables, \code{cotabplot()} is conditioning on variable \emph{levels}.
The plot in Figure \ref{fig:cotabplot} shows the Survival rate among the classes, given
all four level combinations of Sex and Age.

<<cotabplot>>=
cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
<<cotabplot>>
@
\caption{cotabplot for the Titanic data.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\section{Technical Notes on Panel Functions}

All strucplot components (including the panel functions) 
are parameterized through the use of generating functions, that is,
functions with all parameters in their signature returning a function
to be called by \code{strucplot()}. This returned function has access
to all parameters passed to the generating function because in {\sf
  R}, nested functions are lexically scoped. The idea is to use module-specific `code
factories' that generate parameterized workhorses rather than
struggling with monolithic functions with endless parameter lists. As
a simple example, consider \code{struc\_mosaic()} and \code{struc\_assoc},
the generating functions for the mosaic and the association plots:

\begin{smallexample}
struc_mosaic <- function(zero_size = 0.5)
  function(residuals, observed, expected = NULL, spacing, gp, split_vertical) {
  \dots
  }

struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
                        yspace = unit(0.5, "lines"), xscale = 0.9)
  function(residuals, observed = NULL, expected, spacing, gp, split_vertical) {
  \dots
  }
\end{smallexample}

\noindent Both functions have specific parameters in their signature,
but both return a function with a standardized interface that
subsequently is called by \code{strucplot} to produce the actual
plot. As can be seen, the only information \code{strucplot()} passes
to the panel functions are residuals, observed and expected
values, a structure containing the between-tile spacing information,
an object of class \code{"gpar"} for the tiles' graphical parameters,
and a logical vector with the splitting information for each
dimension. The workhorse function, of course, is free to use only
part or all of this information. High-level interface functions such
as \code{mosaic()} and \code{assoc{}} call \code{strucplot()} by
specifying the corresponding function along with the specific
parameters. Now, there a several ways of doing the same thing, as
illustrated by the following equivalent calls:

<<panelfuns>>=
strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
@ 
<<panelfuns2,eval=FALSE>>=
strucplot(Titanic, panel = myworkhorse)
@

\noindent where \code{myworkhorse} could be, e.g., a function with
hard-wired parameters, but with an interface complying with the
workhorse signature. To enable \code{strucplot()} to distinguish
between generating functions and workhorse functions, the generating
functions must be of class \code{"panel\_generator"}.

All panel functions are supposed to produce conditional hierarchical
plots by the means of nested viewports, corresponding to the provided splitting
information. Thus, at the end of the plotting, each tile is associated with
a particular viewport. Each of those viewports has to be
conventionally named, enabling other strucplot modules, in particular the
labeling functions, to access specific tiles after they have been
plotted. Note that the viewport tree is popped by default. The
following example shows how to access parts of the plot after it has
been drawn (see Figure \ref{fig:afterplot}):

<<viewportnames>>=
mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
current.vpTree() ## output not shown

seekViewport("cell:Sex=Male")
grid.rect(gp = gpar(col = "red", lwd = 4))

seekViewport("cell:Sex=Male,Class=Crew")
grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
@

@@

\noindent The naming convention for the viewports is:
\code{cell..\emph{[Variable 1]}.\emph{[Level 1]}..\emph{[Variable 2]}.\emph{[Level 2]}} \dots. Clearly, these names depend on the splitting.

In addition to the viewports, the main graphical elements get names
following a similar construction method. This allows to change
graphical parameters of plot elements after the plotting (see Figure \ref{fig:changeplot}):

<<changeplot>>=
assoc(Titanic)
getNames() ## output not shown
grid.edit("rect:Class=1st,Age=Adult,Sex=Female,Survived=Yes", 
            gp = gpar(fill = "red"))
@

>>>>

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
\newcommand{\data}[1]{`\texttt{#1}'}

%% \code without `-' ligatures
\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
\def\restorehyphenation{\hyphenchar\font=`-}
{\catcode`\-=\active%
  \global\def\code{\bgroup%
    \catcode`\-=\active \let-\codedash%
    \Rd@code}}
\def\codedash{-\discretionary{}{}{}}
\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}

\newcommand{\codefun}[1]{\code{#1()}}

@@

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
data(HairEyeColor)
@

\title{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\email{David.Meyer@R-project.org}
}
\maketitle
\sloppy

The strucplot framework in the {\sf R} package \pkg{vcd} visualizes
multi-way contingency tables and integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into row and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in similar
ways than a two-dimensional table.
This principle defines a class of conditional displays with still many parameters
left such as:

\begin{itemize}
\item the content of the tiles
\item the split direction for each dimension
\item the graphical parameters of the tiles' content
\item the spacing between the tiles
\item the labeling of the tiles
\end{itemize}

@@

The strucplot framework is highly modularized; Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. The workhorse functions (\code{\emph{struc\_foo()}}, 
\code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
directly produce graphical output, the parameter functions
(\code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
graphical parameters used by the others. The \code{\emph{struc\_foo()}} functions
implement the core functionality, creating the tiles and their
content. On the second level, a suitable combination of these functions
is passed as ``hyperparameters'' to \codefun{strucplot}.
These central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several high-level functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
conveniently interface \codefun{strucplot} through sensible parameter defaults
and support for model formulas. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{pairs.table} and
\codefun{cotabplot}) combining single plots of the strucplot 
framework into more complex displays.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{struc}
    \caption{Components of the strucplot framework.}
    \label{fig:struc}
  \end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
<<vcdlayout,fig=TRUE,echo=FALSE,width=6,height=5>>=
    pushViewport(vcd:::vcdViewport(legend = T, mar =4))
    seekViewport("plot")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("plot", gp = gpar(fontsize = 30))
    seekViewport("legend")
    grid.text("legend", rot = 90, gp = gpar(fontsize = 30))
    grid.rect(gp = gpar(lwd = 3))
    seekViewport("pval")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("pval", gp = gpar(fontsize = 30))
    seekViewport("margin_top")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_top", gp = gpar(fontsize = 30))
    seekViewport("margin_bottom")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_bottom", gp = gpar(fontsize = 30))
    seekViewport("margin_right")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_right", rot = 90, gp = gpar(fontsize = 30))
    seekViewport("margin_left")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_left", rot = 90, gp = gpar(fontsize = 30))
    seekViewport("corner_top_left")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[A]", gp = gpar(fontsize = 30))
    seekViewport("corner_top_right")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[B]", gp = gpar(fontsize = 30))
    seekViewport("corner_bottom_left")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[C]", gp = gpar(fontsize = 30))
    seekViewport("corner_bottom_right")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[D]", gp = gpar(fontsize = 30))
@ 
\caption{Viewport layout for strucplot displays with their names. [A] =
  ``corner\_top\_left'', [B] = ``corner\_top\_right'', 
  [C] = ``corner\_bottom\_left'', [D] = ``corner\_bottom\_right''.}
\label{fig:layout}
\end{center}
\end{figure}

\section{Mosaic, Association, and Sieve Plots}

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (sex, i.e. gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base {\sf R}, but allowing the specification of split directions):

<<structable,results=verbatim>>=
(hec <- structable(aperm(HairEyeColor)))
@

Let us first visualize the contingency table by the means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. Figure \ref{fig:observed} depicts 
the effect of

<<Observed>>=
mosaic(hec)
@

\noindent equivalent to

<<Observed2>>=
mosaic(~ Sex + Eye + Hair, data = HairEyeColor)
@

\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent The small bullets indicate zero entries in the corresponding cell.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

<<Expected>>=
mosaic(hec, type = "expected")
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent In order to compare observed and expected values,
a sieve plot \citep{vcd:riedwyl+schuepbach:1994} 
could be used (see Figure \ref{fig:sieve}):

<<sieve>>=
sieve(hec)
@

\begin{figure}[h]
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE>>=
<<sieve>>
@
\caption{Sieve plot for the \data{HairEyeColor} data visualizing simultaneously
  observed and expected values.}
\label{fig:sieve}
\end{center}
\end{figure}

\noindent Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base {\sf R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure \ref{fig:residuals}):

<<Residuals>>=
assoc(hec, compress = FALSE)
@

\begin{figure}[p]
\begin{center}
<<Residualsfig,fig=TRUE,echo=FALSE>>=
<<Residuals>>
@
\caption{Association plot for the \data{HairEyeColor} data.}
\label{fig:residuals}
\end{center}
\end{figure}

@@

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one).

<<split1>>=
mosaic(hec, split_vertical = c(TRUE, TRUE, FALSE))
@

\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

<<split2>>=
mosaic(hec, direction = c("v","h","v"))
@

\begin{figure}[p]
\begin{center}
<<splitfig,fig=TRUE,echo=FALSE>>=
<<split1>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data---alternative splitting.}
\label{fig:split}
\end{center}
\end{figure}

By a suitable combination of splitting, spacing, and
labeling settings, the funtions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
it was almost trivial to implement doubledecker
plots---\codefun{doubledecker} is really just a wrapper for 
\codefun{mosaic}, setting the right defaults. 
Figure \ref{fig:doubledecker} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (`Survived') by Age, given
Sex, given Class. It is created by:

@@

\begin{figure}[h]
\begin{center}
<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
<<doubledecker1>>
@
\caption{Doubledecker plot for the \data{Titanic} data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

\section{Conditional and partial views}

So far, we have visualized full tables. For objects of class \code{table},
conditioning on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is simply done by indexing.
However, subsetting \code{"structable"} objects is more restrictive 
because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
conditioning is only possible ``outside-in'':

<<subsetting,results=verbatim>>=
hec
hec[["Male",]]
hec[[c("Male", "Brown"),]]
hec[["Male","Green"]]
@

\noindent Now, there are several ways for visualizing 
conditional independence structures. The ``brute force'' method is to
draw separate plots for the strata. The following example compares 
the association between hair and eye color, given gender, by using
subsetting on the flat table and the \pkg{grid} package's viewport 
framework to visualize the two groups besides each other:

pushViewport(viewport(layout.pos.col = 1))
mosaic(hec[["Male"]], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(hec[["Female"]], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=5,width=12>>=
<<Variables>>
@
\caption{Distribution of hair and eye color, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent Note the use of the \code{margins} argument: it takes a
vector with up to four values whose unnamed components are recycled,
but ``overruled'' by the named arguments. Thus, in the example, only
the top margin is set to 2 lines, and all other to 0.
This idea applies to almost all
vectorized arguments in the strucplot framework (with
\code{split\_vertical} as a prominent exception).

Since mosaic displays are ``conditional plots'' by definition, we can
also use one single mosaic for stratified plots.
The formula interface of \codefun{mosaic} allows the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

<<Conditioning>>=
mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE)
@

\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}

\noindent The effect of using this kind of formula is that
conditioning variables are permuted before the conditioned variables
in the table, and that \codefun{spacing\_conditional} is used as
default to better
distinguish conditioning from conditioned dimensions.
This spacing uses equal space between tiles of conditioned variables,
and increasing space between tiles of conditioning variables.
Note, however, that the plots in the ``pseudo-strata'' are
distorted since they are not corrected for the marginal
distribution(s) of the conditioning variables.

The \codefun{cotabplot} function does a much better job on this task:
it arranges stratified strucplot displays 
in a lattice-like layout, conditioning on variable \emph{levels}.
The plot in Figure \ref{fig:cotabplot} shows Hair and Eye color, given Sex:

<<cotabplot>>=
cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=4,width=7>>=
<<cotabplot>>
@
\caption{Conditional tabular plot for the \data{HairEyeColor} data.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent The \code{labeling\_args} argument ensures that `Eye' levels
are abbreviated (See the vignette on labeling for detailed information).

Another high-level function for visualizing conditional independence
models is \codefun{pairs.table} (\codefun{pairs.structable}), 
that is, the {\sf S3}-method for objects
of class \code{"table"} (\code{"structable"}) of the \codefun{pairs} generic. 
In contrast to \codefun{cotabplot} which is conditioning on variable \emph{levels},
\codefun{pairs.table} is conditioning on the variables themselves, creating
partial views of the table. The function produces a matrix 
having strucplot displays in the off-diagonal cells, and
the variable names (with, optionally, univariate statistics) in the diagonal cells. 
Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
independence in the lower triangle,
association plots for the same in the upper triangle, and bar charts in the diagonal.

<<pairs>>=
pairs(hec, upper_panel = pairs_assoc, space = 0.3, 
      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<pairsfig,fig=TRUE,echo=FALSE>>=
<<pairs>>
@
\caption{Pairs plot for the \data{HairEyeColor} data.}
\label{fig:pairs}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent In plots produced by \codefun{pairs.table}, 
each cell's row and column define two variables $X$ and $Y$ used for the
specification of four different types of independence: `\code{pairwise}',
`\code{total}', `\code{conditional}' and `\code{joint}'. The pairwise mosaic
matrix shows bivariate marginal relations between $X$ and $Y$, collapsed over all
other variables. The total independence mosaic matrix shows mosaic
plots for mutual independence, i.e., for marginal and conditional independence
among all pairs of variables. The conditional
independence mosaic matrix shows mosaic plots for marginal
independence of $X$ and $Y$, given all other variables. The joint independence
mosaic matrix shows mosaic plots for joint independence of all
pairs ($X$, $Y$) of variables from the others.

Since the matrix is symmetric, the upper and lower parts can
independently be used to display different 
types of independence models, or different strucplots displays
(mosaic, association, or sieve plots).
The available core functions (\codefun{pairs\_assoc}, 
\codefun{pairs\_mosaic}, and \codefun{pairs\_sieve}) 
are just simple wrappers to \codefun{assoc}, 
\codefun{mosaic}, and \codefun{sieve}, respectively. 
Obviously, seeing patterns in strucplot matrices 
becomes increasingly difficult with
higher dimensionality. Typically, therefore, this plot is used with a suitable 
residual-based shading (described in a separate vignette).

\section{Interactive plot modifications}

All strucplot core functions are supposed to produce conditional hierarchical
plots by the means of nested viewports, corresponding to the provided splitting
information. Thus, at the end of the plotting, each tile is associated with
a particular viewport. Each of those viewports has to be
conventionally named, enabling other strucplot modules, in particular the
labeling functions, to access specific tiles after they have been
plotted. The naming convention for the viewports is:

\begin{center}
\code{cell:\emph{[Variable 1]}=\emph{[Level 1]},\emph{[Variable 2]}=\emph{[Level 2]}} \dots
\end{center}

\noindent Clearly, these names depend on the splitting.
The following example shows how to access parts of the plot after it has
been drawn (see Figure \ref{fig:afterplot}):

<<viewportnames>>=
mosaic(~ Hair + Eye, data = hec, pop = FALSE)

seekViewport("cell:Hair=Blond")
grid.rect(gp = gpar(col = "red", lwd = 4))

seekViewport("cell:Hair=Blond,Eye=Blue")
grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
@

\noindent Note that the viewport tree is removed by
default. Therefore, the \texttt{pop} argument has to be set to
\texttt{FALSE} when viewports shall be accessed.

@@

In addition to the viewports, the main graphical elements get names
following a similar construction method. This allows to change
graphical parameters of plot elements \emph{after} 
the plotting (see Figure \ref{fig:changeplot}):

<<changeplot,results=verbatim>>=
assoc(Eye ~ Hair, data = hec, pop = FALSE)
getNames()[1:6]
grid.edit("rect:Hair=Blond,Eye=Blue", gp = gpar(fill = "red"))
@

%% code-chunk reuse does not work with parameter changing
\begin{figure}[h]
\begin{center}
<<changeplotfig,fig=TRUE,echo=FALSE>>=
x = tab = margin.table(HairEyeColor, 1:2)
x[] = "light gray"
x["Blond","Blue"] = "Red"
assoc(tab, gp = gpar(fill = x))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}

\bibliographystyle{apalike}
\bibliography{vcd}
